---
title: "Police Use of Force and County Diversity in Washington State"
author: "Aman Sarosh, Brandon Lay, Gabriel Aritchita"
format:
  dashboard:
    theme: cosmo
    embed-resources: true
---

```{r}
#| message: false
#| warning: false
#| echo: false
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
library(tidyr)

# Load data
diversityindex_df <- read.csv("data_clean/Counties.csv", skip = 1)
rate_of_force_data_df <- read.csv("data_clean/Rate_of_Force_Data_R_1771113615253.csv")

# Diversity data
diversityindex_df <- diversityindex_df |>
  filter(State.Name == "WA") |>
  mutate(County = str_remove(County_Name, " County"))

# Aggregate force data by county
rate_of_force_data_df <- rate_of_force_data_df |>
  group_by(County) |>
  summarise(
    Calls_Service = sum(Calls.for.Service, na.rm = TRUE),
    Use_Force = sum(Uses.of.Force, na.rm = TRUE)
  )

# Join datasets
force_by_diversity_df <- diversityindex_df |>
  left_join(rate_of_force_data_df, by = "County") |>
  select(County, TotPop10, PWhite10, PBlack10, PHisp10, PAPI10,
         POther10, E10, Calls_Service, Use_Force) |>
  mutate(Force_Per_100k = (Use_Force / TotPop10) * 100000) |>
  drop_na()
```

# Overview

## Row {height="25%"}

```{r}
#| content: valuebox
#| title: "Counties Analyzed"
list(
  icon = "map",
  color = "primary",
  value = nrow(force_by_diversity_df)
)
```

```{r}
#| content: valuebox
#| title: "Total Use of Force Incidents"
list(
  icon = "shield-exclamation",
  color = "warning",
  value = scales::comma(sum(force_by_diversity_df$Use_Force))
)
```

```{r}
#| content: valuebox
#| title: "Avg Force per 100k Residents"
list(
  icon = "person",
  color = "danger",
  value = round(mean(force_by_diversity_df$Force_Per_100k), 1)
)
```

## Row {height="75%"}

```{r}
#| title: "Use of Force per 100k Residents by County"
force_by_diversity_df |>
  arrange(desc(Force_Per_100k)) |>
  ggplot(aes(x = reorder(County, Force_Per_100k), y = Force_Per_100k, fill = Force_Per_100k)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#7fbfff", high = "#c0392b") +
  coord_flip() +
  labs(
    x = "County",
    y = "Use of Force per 100,000 Residents"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank()
  )
```

# Data Table

## Row

```{r}
#| title: "County-Level Summary"
force_by_diversity_df |>
  arrange(desc(Force_Per_100k)) |>
  select(County, TotPop10, E10, Use_Force, Force_Per_100k) |>
  rename(
    "Population (2010)" = TotPop10,
    "Diversity Index" = E10,
    "Uses of Force" = Use_Force,
    "Force per 100k" = Force_Per_100k
  ) |>
  mutate(
    "Population (2010)" = scales::comma(`Population (2010)`),
    "Diversity Index" = round(`Diversity Index`, 3),
    "Force per 100k" = round(`Force per 100k`, 1)
  ) |>
  knitr::kable()
```

# Research Questions

## Row

**1.** Is there a relationship between a county's diversity index and its rate of police use of force per 100,000 residents?

**2.** Which racial/ethnic population proportions are most strongly associated with elevated use-of-force rates across Washington State counties?

**3.** Do counties with higher call-for-service volumes show proportionally higher use-of-force incidents, or do some counties apply force disproportionately relative to call volume?