---
title: "Police Use of Force and County Diversity in Washington State"
author: "Aman Sarosh, Brandon Lay, Gabriel Aritchita"
format: 
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| message: false
#| warning: false
#| echo: false
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
library(tidyr)
```

# Section I: First Dataset - WADEPS Police Use of Force

## Data Source

The first dataset comes from **WADEPS** (Washington State Data Exchange
for Public Safety), which is run by Washington State University under a
grant from the Washington State Attorney General's Office. State law
enacted in 2021 requires all police agencies in
Washington to report use of force incidents monthly to this centralized
database.

This source is highly reliable because:

- It is legally mandated reporting with oversight from the Attorney General
- Data is collected and managed by university researchers with government accountability
- The database launched in June 2022 and can be verified through FOIA requests, news coverage, and direct agency reporting

**Source Link:** <https://wadeps.org/home/dashboard/>

## What It Represents

This dataset documents police use of force incidents across Washington
State at the agency level. Each row represents a police agency (city
police departments, county sheriffs, etc.) and includes:

- Total calls for service received by that agency
- Total reported uses of force by that agency

This data is directly relevant to our research question: **Do counties
with different racial and ethnic compositions show different patterns in
police use of force?** By linking this use of force data with
demographic information, we can investigate potential correlations
between community diversity and police force incidents.


## Data Overview

```{r load-force-data}
#| echo: true
rate_of_force_data_df <- read.csv("data_clean/Rate_of_Force_Data_R_1771113615253.csv")
glimpse(rate_of_force_data_df)
```

**Key Characteristics:**

- **87 rows** representing different law enforcement agencies
- **5 columns**: Agency, Agency Type, County, Calls for Service, Uses of Force
- Coverage: 25 out of 39 Washington counties


## Join Plan

We will join this dataset with our diversity data using **County** as
the key variable. Some data preparation will be needed:

- County names must match exactly between datasets
- Multiple agencies within the same county need to be aggregated
- Remove " County" suffix from diversity dataset county names

## Data Issues

**Missing Coverage:** Only 25 of Washington's 39 counties are
represented. This limitation likely stems from rural counties lacking
dedicated police departments or having incomplete reporting compliance.

**No Missing Values:** Within the reported data, all values are present
and complete.

**Implication:** Our analysis will only cover counties with active law
enforcement agencies that reported to WADEPS, potentially excluding more
rural areas with different policing dynamics.

# Section II: Second Dataset - County Diversity Index


## Data Source

The second dataset comes from the **Diversity and Disparities Project**
at Brown University, a research initiative examining
racial diversity and disparities across American communities. This
project collaborates directly with U.S. Census data to provide
peer-reviewed demographic analysis.

This source is reliable because:

- Academic research from Brown University with rigorous standards
- Based on official U.S. Census Bureau data
- Peer-reviewed and widely cited in demographic research

**Source Link:**
<https://s4.ad.brown.edu/projects/diversity/Data/Download4.htm>

## What It Represents

This dataset measures racial and ethnic diversity for every U.S. county
from 1980-2010.

For each county, the data includes:

- Total population counts by census year
- Racial/ethnic composition percentages (White, Black, Hispanic, Asian/Pacific Islander, Other)

**Relevance to Research:** This demographic context is required for
understanding whether police use of force correlates with community
diversity levels or specific racial compositions.


## Data Overview

```{r load-diversity-data}
#| echo: true
diversityindex_df <- read.csv("data_clean/Counties.csv", skip = 1)
glimpse(diversityindex_df)
```

**Key Characteristics:**

- **3,143 rows** (all U.S. counties)
- **64 columns** including population, racial percentages, entropy scores
- Census years: 1980, 1990, 2000, 2010
- We will filter to Washington State only (39 counties)


## Join Plan

Join on **County Name** after:

- Filtering to only Washington State counties (`State.Name == "WA"`)
- Removing " County" suffix to match format in WADEPS data
- Using 2010 census data (most recent in this dataset)

## Data Issues

**Missing Values:** The dataset contains scattered missing values and
NAs, particularly in smaller counties or specific demographic categories.
These will be handled through filtering during the join process.

**Census Timing:** Using 2010 Census data with 2022-2025 police data
introduces a temporal mismatch. County demographics have changed over
12-15 years, which may affect analysis accuracy.

**Geographic Aggregation:** County-level data masks within-county
variation, particularly in large counties like King County (Seattle area)
where diversity varies significantly by neighborhood.

# Section III: Joining the Data

## Data Join Process

```{r data-join}
#| echo: true
#| message: false
#| warning: false

# Prepare diversity data - filter to WA and clean county names
diversityindex_df <- diversityindex_df |>
  filter(State.Name == "WA") |>
  mutate(County = str_remove(County_Name, " County"))

# Aggregate force data by county (multiple agencies per county)
rate_of_force_data_df <- rate_of_force_data_df |>
  group_by(County) |>
  summarise(
    Calls_Service = sum(Calls.for.Service, na.rm = TRUE),
    Use_Force = sum(Uses.of.Force, na.rm = TRUE)
  )

# Join datasets
force_by_diversity_df <- diversityindex_df |>
  left_join(rate_of_force_data_df, by = "County")

# Select relevant columns
force_by_diversity_df <- force_by_diversity_df |>
  select(County, TotPop10, PWhite10, PBlack10, PHisp10, PAPI10, 
         POther10, E10, Calls_Service, Use_Force) |>
  mutate(Force_Per_100k = (Use_Force / TotPop10) * 100000) |>
  drop_na()

# Display joined data
glimpse(force_by_diversity_df)
```

### Join Methodology

**Changes Made:**

1. **Filtered** diversity data to Washington State only
2. **Removed** " County" suffix from county names in diversity data
3. **Aggregated** multiple agencies per county (e.g., King County has
   Seattle PD, Bellevue PD, etc.)
4. **Joined** using left join on County name
5. **Selected** only relevant demographic and force variables
6. **Created** normalized metric: Force_Per_100k (accounts for
   population differences)
7. **Removed** rows with missing data using `drop_na()`

### Final Dataset Structure

```{r show-final-structure}
#| echo: false
kable(head(force_by_diversity_df, 5), 
      caption = "First 5 rows of joined dataset",
      digits = 2)
```

**Final Variables:**

- `County`: County name
- `TotPop10`: Total 2010 population
- `PWhite10`: % White population
- `PBlack10`: % Black population
- `PHisp10`: % Hispanic population
- `PAPI10`: % Asian/Pacific Islander
- `POther10`: % Other races
- `E10`: Entropy diversity index
- `Calls_Service`: Total calls for service
- `Use_Force`: Total uses of force
- `Force_Per_100k`: Uses of force per 100,000 residents


## Data Documentation

### Codebook Links

- **Diversity Data:**
  <https://s4.ad.brown.edu/projects/diversity/DiversityPages2/Default.aspx>
- **WADEPS Data:** <https://wadeps.org/home/dashboard/> (WADEPS
  provides a dashboard along with CSV)

### Most Relevant Variables

**Primary Outcome Variables:**

- `Force_Per_100k`: Normalized rate of force usage, our main dependent
  variable
- `Use_Force`: Raw count of force incidents

**Key Explanatory Variables:**

- `E10`: Entropy diversity index - provides a single comprehensive
  diversity score
- `PWhite10`, `PBlack10`, `PHisp10`, `PAPI10`, `POther10`: Racial
  composition percentages for granular analysis

**Control Variables:**

- `TotPop10`: Population size for context
- `Calls_Service`: Police activity level (distinguishes high force rates
  from high activity)

The most meaningful analyses will examine relationships between
`Force_Per_100k` and both the overall diversity measure (`E10`) and
individual racial composition variables to identify potential disparities.

### Potential Biases and Human Rights Principles

Our dataset contains several important limitations:

**Accountability & Transparency:** The force data only includes agencies
that reported to WADEPS. Departments with compliance issues or serious
problems may be systematically excluded, weakening accountability. This
represents only 25 of 39 counties.

**Fairness & Non-Discrimination:** The PAPI10 variable aggregates all
Asian and Pacific Islander identities into one category, masking
important differences between diverse communities (e.g., East Asian,
Southeast Asian, Pacific Islander populations have distinct experiences).
This aggregation undermines equitable analysis.

**Transparency & Explainability:** We lack crucial variables including
officer demographics, socioeconomic factors (poverty, education,
employment), crime rates, and police department policies. These omissions
limit our ability to explain observed disparities.

**Safety & Security:** Using 2010 Census data with 2022-2025 police data
creates a temporal mismatch. Population demographics have shifted
significantly over 12-15 years, particularly in rapidly growing counties
like Clark and Snohomish.

**Privacy:** County-level aggregation protects individual privacy but
reduces analytical precision. Within-county variation is masked,
particularly problematic in large, diverse counties like King County.

**Human Control & Values:** The dataset lacks community perspectives,
civilian oversight data, or input from affected populations. This
top-down approach to data collection weakens community voice in
understanding police-community relations.

**Professional Responsibility:** As researchers, we have a responsibility
to clearly communicate these limitations and avoid overstating
conclusions about systemic policing disparities. Our analysis can
identify correlations but cannot establish causation or provide complete
explanations for observed patterns.

# Section IV: Data Wrangling

In this section, we perform three data cleaning and exploration actions on our combined dataset to prepare it for analysis.

## Action 1: Filter to Washington State and Clean County Names

We filtered the diversity dataset to only include Washington counties and removed the " County" suffix from county names to enable proper joining.
```{r action-1-filter-clean}
#| echo: true
#| eval: false

# This was done in the join section
diversityindex_df <- diversityindex_df |>
  filter(State.Name == "WA") |>
  mutate(County = str_remove(County_Name, " County"))
```

**Rationale:** The diversity dataset contains all U.S. counties, but we only need Washington State data. Removing " County" from names ensures the county identifiers match between our two datasets for successful joining.

## Action 2: Aggregate Multiple Agencies per County

Since multiple law enforcement agencies can operate within the same county, we grouped and summarized the force data by county to get total counts.
```{r action-2-aggregate}
#| echo: true
#| eval: false

# This was done in the join section
rate_of_force_data_df <- rate_of_force_data_df |>
  group_by(County) |>
  summarise(
    Calls_Service = sum(Calls.for.Service, na.rm = TRUE),
    Use_Force = sum(Uses.of.Force, na.rm = TRUE)
  )
```

**Rationale:** Depending on county size, there may be 1-4 agencies within it (e.g., King County has Seattle PD, Bellevue PD, sheriff's office, etc.). We combined these to get county-level totals that can be compared with county-level demographics.

## Action 3: Create Normalized Force Rate Variable

We created a new column that calculates the rate of force incidents per 100,000 residents to account for population differences between counties.
```{r action-3-normalize}
#| echo: true
#| eval: false

# This was done in the join section
force_by_diversity_df <- force_by_diversity_df |>
  mutate(Force_Per_100k = (Use_Force / TotPop10) * 100000) |>
  drop_na()
```

**Rationale:** Counties with larger populations will naturally have more calls for service and uses of force. Creating a normalized rate (per 100,000 people) allows fair comparison between counties of different sizes. We also removed any rows with missing data to ensure data quality.

## Summary of Cleaned Dataset
```{r final-summary}
#| echo: true

# Show structure of final dataset
glimpse(force_by_diversity_df)

# Show summary statistics
summary(force_by_diversity_df$Force_Per_100k)
```

Our final cleaned dataset contains **25 counties** with complete data on demographics and police use of force. The Force_Per_100k variable will be our primary metric for comparing force rates across counties with different diversity levels.

# Section V: Group Information

## Project Overview

**Topic:** Police Use of Force and County Demographics in Washington State

**Lab Section:** INFO 201 B

**Teaching Assistant:** Yashita Tanwar

**Research Question:** This project investigates whether patterns in
police use of force correlate with racial and ethnic diversity across
Washington State counties. We examine both overall diversity and specific demographic compositions to understand
potential disparities in law enforcement practices.

## Team Members

### Aman Sarosh

**Affiliation:** The Information School, University of Washington - Seattle

**Email:** asarosh@uw.edu

**Project Interest:** I am interested in examining how demographic
factors influence public safety outcomes and whether data can reveal
systemic patterns in law enforcement that warrant policy attention.

**Informatics Background:** One of my interest in Informatics centers on using
data analysis to address questions and inform evidence-based policy decisions that promote equity.

"Study, Design, and Develop information technology for the good of people, organizations, and society."

### Brandon Lay

**Affiliation:** The Information School, University of Washington - Seattle

**Email:** bndonlay@uw.edu

**Project Interest:** I am interested in understanding the intersection
of community diversity and policing practices, particularly how data
transparency can support accountability in law enforcement.

**Informatics Background:** Im drawn to informatics because it tackles solving issues by utilizing hidden solutions within data. Within my career specifically Im interested in analytics as it matches my aspirations within informatics

### Gabriel Aritchita

**Affiliation:** The Information School, University of Washington - Seattle

**Email:** gabe1124@uw.edu

**Project Interest:** I am interested in exploring whether and how
racial composition of communities affects their interactions with law
enforcement, with the goal of identifying areas for reform.

**Informatics Background:** I am passionate about using data science to investigate questions of fairness and equity.

## Coding Notes

### Tools and Resources Used

- **R and RStudio:** Core programming environment
- **dplyr package:** Data manipulation and joining
- **ggplot2 package:** Data visualization
- **tidyverse package:** Data science tools
- **Quarto:** Report generation and reproducible research
- **readr package:** Data import functions 
- **stringr package:** String manipulation for cleaning county names
- **knitr package:** Creating tables in reports
- **Google:** General Syntax