---
title: "Police Use of Force by Ethnic Population"
author: "Aman Sarosh, Brandon Lay, Gabriel Aritchita"
format: html
theme: cosmo
embed-resources: true
editor: 
  markdown: 
    wrap: 72
---

## Data Sources

<https://wadeps.org/home/dashboard/>

<https://s4.ad.brown.edu/projects/diversity/Data/Download4.htm>

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(stringr)

source("R/load_data.R")
```

# Part I: Find the Data

## First Dataset - WADEPS Police Use of Force

### Data Source

The first dataset comes from WADEPS, the Washington State Data Exchange
for Public Safety. WADEPS is run by Washington State University under a
grant from the Attorney General's Office. State law from 2021 requires
all police agencies in Washington to report use of force incidents
monthly.

This is reliable because it's legally mandated and managed by university
researchers with government oversight. The database launched in June
2025 and you can verify it through news coverage and the agencies are
real Washington police departments.

### What It Represents

This data shows police use of force incidents across Washington State.
Each row is a police agency showing how many calls for service they got
and within that department how many repoted use of force cases are tied
with it.

This connects to demographics because we want to see if counties with
different racial compositions have different patterns of police force.

### Join Plan

We'll join on county name. Both datasets should have county names but we
might need to clean them up to make sure they match exactly.

### Data Overview

```{r}
glimpse(wadeps_df)

head(wadeps_df, 8)
```

The dataset has 86 rows for different agencies and 5 columns: Agency,
Agency Type, County, Calls for Service, and Uses of Force.

### Data Issues

Looking at the dataset as a whole, there are no missing values and 25 of
39 counties are represented so the dataset purely encompasses counties
that have reported documents of their calls for service and use of
force. This shortness of counties could be rooted in the more rural
counties that may not have a dedicated police department attactched to
the lcoation.

## Second Dataset - County Diversity Index

### Data Source

The source is called Diversity and Disparities, it's a project within
Brown University that provides research that examines as their title
suggests the diversity and disparties within American society. All of
their research is independent as well as peer reviewed in collaboration
with census data. They themselves sort and collect the data alongside
census and I believe its reiable due to it being presented with the
connection of the prestigious college Brown University.

### What It Represents

The data represents how racially diverse each U.S county is from 1980 -
2010 using an entropy scope. This information is relevant to the first
dataset because it provides demographic context for counties where
police use of force incidents occured. By linking and comparing the two,
you can analyze whether a specific county's level of racial diversity is
related to how often force is used.

### Join Plan

We'll join on county name. Need to make sure the county names match
between the two datasets and filter to just Washington counties.

### Data Overview

```{r}
glimpse(diversity_df)

head(diversity_df, 8)
```

The dataset has 64 columns including county identifiers, metro area
information, and census data that encompass population count racial
percentage and an entropy scope. Theres 3143 rows encompassing every
county within the U.S. An example of the data would be, Autauga County,
Alabama had a 2010 population of 54,571 with a ratio of 77% white and
17.6% black, and its entropy scope increasing from 38.7 in 1980 to 44.2
in 2010.

### Data Issues

Some blank/missing data sets as well as a few N/As here and there, too
many to count but from the looks of it they look scattered and would
require some forms of clenaing before fully implementing it into a join
function between our two datasets.

# Part II: Data Join

```{r}
library(dplyr)
library(tidyverse)

diversityindex_df <- read.csv("~/Downloads/Counties.csv" , skip = 1)
rate_of_force_data_df <- read.csv("~/Downloads/Rate_of_Force_Data_R_1771113615253.csv")

diversityindex_df <- diversityindex_df |>
  filter(State.Name == "WA") |>
  mutate(County = str_remove(County_Name, " County"))

rate_of_force_data_df <- rate_of_force_data_df |>
  group_by(County) |>
  summarise(Calls_Service = sum(Calls.for.Service),
            Use_Force = sum(Uses.of.Force))
                                  
force_by_diversity_df <- diversityindex_df |>
  left_join(rate_of_force_data_df, by = "County")

force_by_diversity_df <- force_by_diversity_df |>
  select(County, TotPop10, PWhite10, PBlack10, PHisp10, PAPI10, POther10, E10, Calls_Service, Use_Force) |>
  mutate(Force_Per_100k = (Use_Force / TotPop10) * 100000) |>
  drop_na()

write.csv(force_by_diversity_df, "force_by_diversity_df.csv", row.names = FALSE)
                            
```

Changes made before joining: - Defined the two datasets with their
respective names - Started by filtering to only counties within the
state of washington - Found that diversityindex_df utilizes county at
the end of their name so it had to be removed (ex Kings County -\>
Kings) - There are multiple agencies within the same county depending on
its size so they had to be combined to fully encompass the county -
Changed the names of calls for service and use of force to just
Calls_Service and Use_Force - Then defined a new combined dataset with
its respective name and did a left join by the dataset's counties -
Selected only the columns we are going to be using to draw our
comparisons such as race, population, diversity index, calls of service,
and reported use of force - Counties who contain a larger population
will obviously have a larger amount of calls and force so we creaetd a
new colum that tells how many uses of force occurred for every 100,000
people living in that county.

```{r}
glimpse(force_by_diversity_df)
```

Utilizing glimpse we can see that our data has 11 columns those being
County, Total Population, White Population, Black Population, Hispanic
Population, Asian and Pacific Islander Population, Ethnic Diversity
Index Value, Calls of Service, Use of Force, and finally Force Per
100,000 population. For the rows we have a total of 25 rows.

# Part III: Data Documentation

## Variable Descriptions

CODEBOOK Links
<https://s4.ad.brown.edu/projects/diversity/DiversityPages2/Default.aspx>

<https://wadeps.org/home/dashboard/> Note: WADEPS doesn't offer a direct
code book but rather the closest thing is their dashboard that utilizes
the dataset which contains a brief description of what the specific
dataset contains.

## Most Relevant Variables

The most relevant variables are Force_Per_100k and Use_Force as they
directly measure our research question. Calls_Service provides context
by distinguishing counties with genuinely high force rates from those
with simply more police activity. In regards to the demographics,
PWhite10, PBlack10, PHisp10, PAPI10, and POther10 are very essential for
analyzing whether counties with higher minority populations experience
disproportionate force. E10 is particularly useful as a single diversity
score for each county. TotPop10 ensures our comparisons are population
adjusted rather than skewed by county size. The most meaningful analysis
will come from plotting Force_Per_100k against each racial percentage
variable to identify whether any ethnic demographic correlates with
higher use of force across Washington counties.

## Potential Biases and Human Rights Principles

Our dataset contains clear biases that need to be acknowledged. First,
the force data only includes agencies that chose to report, which means
departments with more serious issues may be missing. That weakens
accountability and transparency. The PAPI10 category groups multiple
Asian and Pacific Islanders identities as a group, which hurts fairness
and non-discrimination because it hides important differences. We are
also missing key variables like officer demographics and the economic
factors; it limits transparency and explainability when analying
disparities. Using 2010 Census also raises concerns since county
populations have changed over time. While county level aggregation helps
protect privacy, it reduces detail and masks within county variation.
The dataset also lacks direct community perspectives, which weakens
human control of technology and the promotion of human values.
Personally I believe that as researchers our group has a responsibility
to clearly state these limitations and avoid overstating conclusions
about system policing disparities.

# Part IV: Combined Data Wrangling

## 3 Forms of Cleaning

Note: A lot of this will be a restatement of the previous statement as
both address similar things

```{r}
diversityindex_df <- diversityindex_df |>
  filter(State.Name == "WA") |>
  mutate(County = str_remove(County_Name, " County"))

rate_of_force_data_df <- rate_of_force_data_df |>
  group_by(County) |>
  summarise(Calls_Service = sum(Calls.for.Service),
            Use_Force = sum(Uses.of.Force))

force_by_diversity_df <- force_by_diversity_df |>
  select(County, TotPop10, PWhite10, PBlack10, PHisp10, PAPI10, POther10, E10, Calls_Service, Use_Force) |>
  mutate(Force_Per_100k = (Use_Force / TotPop10) * 100000) |>
  drop_na()

```

## Explanation For Cleaning

The first section mainly addresses removing the state names that arent
Washington as well as making a new column called County that contains
the original county names without the county part to them. We did this
to assist in making sure that the join function works

The second section addresses how multiple counties can contain multipl
agencies. Depending on the size a county may have 1-4 agencies within it
so to help solve that we goruped the counties so the multiple King
Counties become just one King County and that applies to their calls of
service and their use of force.

The third section helps finalize the combined dataset by selecting only
columns that we require and are applicable to the question we are
addressing. Alongside this is a new column created to encompass the rate
at which force is used per 100,000 people in a population. This helps
remove the potentional factors that may contribute to a higher use of
force rate like a higher population count.
